---
title: "DAP test"
author: "Dania Usman"
date: "2024-11-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r}

library(tidyverse)
eng_ed <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-01-23/english_education.csv')


```


# Data Cleaning


```{r}

#selecting useful columns and changing variable names

eng_ed <- eng_ed |> 
  select(town11nm, size_flag, income_flag, university_flag,
         level4qual_residents35_64_2011,
         key_stage_4_attainment_school_year_2012_to_2013,
         highest_level_qualification_achieved_by_age_22_level_6_or_above,
         ) |> 
    mutate(
    size_flag = factor(size_flag),
    income_flag = factor(income_flag),
    uni_flag= factor(university_flag),
    qual_residents=factor(level4qual_residents35_64_2011),
    GCSEs=key_stage_4_attainment_school_year_2012_to_2013,
    college_grad =  highest_level_qualification_achieved_by_age_22_level_6_or_above
   ) 

#removing repeat columns

eng_ed <- eng_ed |> 
  select(town11nm, size_flag, income_flag, uni_flag,qual_residents, 
                 GCSEs, college_grad   )

glimpse(eng_ed)
summary(eng_ed)



```

```{r}


#factor collapse to combine mid and large size towns and cities into one level
#ignored london as other
levels(eng_ed$size_flag)
eng_ed<- eng_ed |>
mutate(size_flag = fct_collapse(size_flag,
Large = c("City", "Medium Towns", "Large Towns", "Outer london BUA", "Inner London BUA", "Not BUA"),
Small = c("Small Towns",  "Other Small BUAs")
)) 



barplot(table(eng_ed$size_flag), las = 2) #distribution of the size_flag
```


# High School Graduation Rates small vlarge towns

Is there a difference in high school graduation rates between small v large towns/cities?
H0: There is no mean difference between the high school graduation rates  between small v large towns/cities
H1: There is a difference  between the high school graduation rates  between small v large towns/cities
Test: Difference between two means

##Conditions

Independence: We can assume the observations of one town to another are independent from each other.


```{r}
small_town <- eng_ed |> 
  filter(size_flag == "Small") #filtered data by small town

large_town <- eng_ed |> 
  filter(size_flag == "Large") #filtered data by large town

hist(small_town$GCSEs)

#look normal

hist(large_town$GCSEs)


#look normal

nrow(small_town)
nrow(large_town)

```


# Power Analysis

```{r}
library(pwr)

mean_small_gcse <- mean(small_town$GCSEs, na.rm = TRUE)
mean_large_gcse <- mean(large_town$GCSEs, na.rm = TRUE)

sd_pooled_gcse <- sqrt((
  (sd(small_town$GCSEs, na.rm = TRUE)^2 + sd(large_town$GCSEs, na.rm = TRUE)^2) / 2
))

# Cohen's d (effect size)
cohens_d_gcse <- abs(mean_small_gcse - mean_large_gcse) / sd_pooled_gcse

# Sample size
n_small <- nrow(small_town)
n_large <- nrow(large_town)

# Power analysis
power_gcse <- pwr.t.test(
  d = cohens_d_gcse,   
  n = min(n_small, n_large), 
  sig.level = 0.05,   
  type = "two.sample" 
)$power

cat("Observed Cohen's d (Effect Size):", round(cohens_d_gcse, 3), "\n")
cat("Sample size (Small Towns):", n_small, "\n")
cat("Sample size (Large Towns):", n_large, "\n")
cat("Computed Power for GCSE Difference Test:", round(power_gcse, 3), "\n")

```


## Test for diff in mean

```{r}
t_test_result <- t.test(small_town$GCSEs, large_town$GCSEs)
cat("T-test Result: \n")
print(t_test_result)
```


Decision: The p-value is less than 0.05, we reject the null hypothesis

Conclusion: We have enough evidence that there is a significant difference in high school graduation rates between small towns and large towns/cities.


# Colleg Graduation Rates small vlarge towns

Is there a difference in college graduation rates between small  v large towns/cities?
H0: There is no difference in college graduation rates between small vs large towns/cities.
H1: There is a difference in college graduation rate between small vs large towns/cities.
Test: Difference between two means

## Conditions
```{r}
hist(small_town$college_grad)
hist(large_town$college_grad)
```

## Power Analysis

```{r}
mean_small_college <- mean(small_town$college_grad, na.rm = TRUE)
mean_large_college <- mean(large_town$college_grad, na.rm = TRUE)

sd_pooled_college <- sqrt((
  (sd(small_town$college_grad, na.rm = TRUE)^2 + sd(large_town$college_grad, na.rm = TRUE)^2) / 2
))

# Cohen's d (effect size)
cohens_d_college <- abs(mean_small_college - mean_large_college) / sd_pooled_college

# Power analysis for two-sample t-test
power_college <- pwr.t.test(
  d = cohens_d_college,   
  n = min(n_small, n_large), 
  sig.level = 0.05,       
  type = "two.sample"     
)$power

cat("Observed Cohen's d (Effect Size):", round(cohens_d_college, 3), "\n")
cat("Sample size (Small Towns):", n_small, "\n")
cat("Sample size (Large Towns):", n_large, "\n")
cat("Computed Power for College Graduation Difference Test:", round(power_college, 3), "\n")


```


## test for diff in means

```{r}
t_test_college_grad <- t.test(small_town$college_grad, large_town$college_grad)
cat("T-test Results for College Graduation Rates: \n")
print(t_test_college_grad)
```

Decision: The p-value is less than 0.05, we reject the null hypothesis

Conclusion: We enough evidence that there is a difference in college graduation rate between small vs large towns/cities.




#Are income and town size assoicated

Independence: 

```{r}
test <- chisq.test(eng_ed$income_flag, eng_ed$size_flag)
cat("Expected counts:", test$expected)
test

```

All counts are greater than 5. 

p value is smaller than 0.05. Income and town size are associated. 


##Power Analysis

```{r}
chisq_test <- chisq.test(eng_ed$size_flag, eng_ed$income_flag)

income_size_table <- table(eng_ed$size_flag, eng_ed$income_flag)

# Calculate Cramér's V
cramers_v <- sqrt(chisq_test$statistic / (sum(income_size_table) * (min(dim(income_size_table)) - 1)))


# Df
df <- chisq_test$parameter

# Total sample size
n_total <- sum(income_size_table)

# Compute power for chi-squared test
power_chisq <- pwr.chisq.test(
  w = cramers_v,  
  N = n_total,  
  df = df,       
  sig.level = 0.05  
)$power

cat("Observed Cramér's V (Effect Size):", round(cramers_v, 3), "\n")
cat("Degrees of Freedom:", df, "\n")
cat("Total Sample Size:", n_total, "\n")
cat("Computed Power for Chi-Squared Test:", round(power_chisq, 3), "\n")
```

# Conditions for linear regression

H0: There is no linear relationship between high school and college completion.
H1: There is a linear relationship between high school and college completion.
Test: Linear Regression

```{r}

#Linearity

ggplot(data = eng_ed, aes(x = GCSEs, y = college_grad)) +
geom_point() +
geom_smooth(method =lm, se =F)


#Independence

mod<- lm(college_grad ~ GCSEs, data =eng_ed)
plot(predict(mod), rstandard(mod))
abline(h=0)
summary(mod)

#constant variance
#normality 
plot(mod) 

```


## Power Analysis

```{r}

r_squared <- summary(mod)$r.squared

# Size (Cohen's f2)
f2 <- r_squared / (1 - r_squared)

n <- nrow(eng_ed) 
num_predictors <- 1

# Power analysis 
power_regression <- pwr.f2.test(
  u = num_predictors, 
  v = n - num_predictors - 1,
  f2 = f2,  
  sig.level = 0.05 
)$power


cat("R-squared (Goodness-of-Fit):", round(r_squared, 3), "\n")
cat("Effect Size (Cohen's f²):", round(f2, 3), "\n")
cat("Sample Size:", n, "\n")
cat("Computed Power for Regression Test:", round(power_regression, 3), "\n")
```


Does having a university in town affect avg proportion of college grad? (diff in mean)

H0:No difference in avg proportion of college grad between univeristy towns and towns with no university
HA: There is a diff. 

Additional Analyses

```{r}
uni_town <- eng_ed |> 
  filter(uni_flag == "University") #filtered data by small town
nrow(uni_town)

no_uni_town <- eng_ed |> 
  filter(uni_flag == "No university") #filtered data by small town
nrow(no_uni_town) 


t.test(uni_town$college_grad, no_uni_town$college_grad)
t.test(uni_town$GCSEs, no_uni_town$GCSEs)

```
p value high. There is no difference in high school and college grad rates between uni towns and non uni towns. 



